---
title: 'Text mining, sentiment analysis, and visualization'
author: "Marcus Fryd"
date: 'created on 22 November 2020 and updated `r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(here)

# For text mining:

library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)

# For text mining in Danish:

library(Sentida)
Sys.setlocale(category = "LC_ALL", locale = "UTF-8")

```

**Note:** for more text analysis, you can fork & work through Casey O’Hara and Jessica Couture’s eco-data-sci workshop (available here https://github.com/oharac/text_workshop)

### Getting the Encomium:
```{r get-document}
encomium_path <- here("data","Anden Bog - Kong Knuts Liv og Gerninger.pdf")
encomium_text <- pdf_text(encomium_path)
```

Some things to notice:

- How cool to extract text out of a PDF! Do you think it will work with any PDF?
- Each row is a page of the PDF (i.e., this is a vector of strings, one for each page)
- The pdf_text() function only sees text that is "selectable"

Example: Just want to get text from a single page (e.g. Page 9)? 
```{r single-page}
encomium_p9 <- encomium_text[9]
encomium_p9
```

See how that compares to the text in the PDF on Page 9. What has pdftools added and where?

From Jessica and Casey's text mining workshop: “pdf_text() returns a vector of strings, one for each page of the pdf. So we can mess with it in tidyverse style, let’s turn it into a dataframe, and keep track of the pages. Then we can use stringr::str_split() to break the pages up into individual lines. Each line of the pdf is concluded with a backslash-n, so split on this. We will also add a line number in addition to the page number."

### Some wrangling:

- Split up pages into separate lines (separated by `\n`) using `stringr::str_split()`
- Unnest into regular columns using `tidyr::unnest()`
- Remove leading/trailing white space with `stringr::str_trim()`

```{r split-lines}
encomium_df <- data.frame(encomium_text) %>% 
  mutate(text_full = str_split(encomium_text, pattern = '\n')) %>% 
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full)) 
head(encomium_df)

# More information: https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html
```

Now each line, on each page, is its own row, with extra starting & trailing spaces removed. 

### Get the tokens (individual words) in tidy format

Use `tidytext::unnest_tokens()` (which pulls from the `tokenizer`) package, to split columns into tokens. We are interested in *words*, so that's the token we'll use:

```{r tokenize}
encomium_tokens <- encomium_df %>% 
  unnest_tokens(word, text_full)
encomium_tokens
# See how this differs from `encomium_df`
# Each word has its own row!
```

Let's count the words!
```{r count-words}
encomium_wc <- encomium_tokens %>% 
  count(word) %>% 
  arrange(-n)
encomium_wc
```

OK...so we notice that a whole bunch of things show up frequently that we might not be interested in ("og", "at", "i", etc.). These are called *stop words*. Let's remove them. 

### Remove stop words:

We will *remove* stop words using a Danish stop word list text file and `tidyr::anti_join()`:
```{r stopwords}
# I will start by loading the stop word list in as a tibble
stop_words_dk <- read_csv("data/stopwordlistDK.txt")
  

# We can now use the anti_join() function to exclude all stop words from our encomium_tokens tibble
encomium_stop <- encomium_tokens %>% 
  anti_join(stop_words_dk) %>% 
  select(-encomium_text)
```

Now we can check the counts again: 
```{r count-words2}
encomium_swc <- encomium_stop %>% 
  count(word) %>% 
  arrange(-n)
encomium_swc
```

What if we want to get rid of all the numbers (non-text) in `encomium_stop`? And what about the declined words like "kongen"?
```{r filter}
# This code will filter out numbers by asking:
# If you convert to as.numeric, is it NA (meaning those words)?
# If it IS NA (is.na), then keep it (so all words are kept)
# Anything that is converted to a number is removed
# "knut" and "knuts" are merged into "knut"
# "kong", "konge", and "kongen" are merged into "konge"
# The merging was done with the help of a friend

encomium_no_numeric <- encomium_stop %>% 
  filter(is.na(as.numeric(word))) 

encomium_no_numeric <- encomium_no_numeric %>% 
  count(word) %>% 
    mutate(word=trimws(as.character(word)),
    group=case_when(word %in% c("knut", "knuts") ~ "knut",
      word %in% c("kong", "konge", "kongen") ~ "konge",
      TRUE ~ word)) %>%
  group_by(group) %>%
  summarise(n = sum(as.numeric(n)), .groups = "drop") %>% 
  arrange(-n) 

encomium_no_numeric
```

### A word cloud of encomium words (non-numeric)

See more: https://cran.r-project.org/web/packages/ggwordcloud/vignettes/ggwordcloud.html

```{r wordcloud-prep}
# There are some 2300 unique words 
length(unique(encomium_no_numeric$word))

# We probably don't want to include them all in a word cloud. Let's filter to only include the top 100 most frequent?
encomium_top100 <- encomium_no_numeric %>% 
  head(100)
encomium_top100
```

```{r wordcloud}
# Let's now actually make the word cloud. We will do so via the tidytext package's geom_text_wordcloud() function:

encomium_cloud <- ggplot(data = encomium_top100, aes(label = group)) +
  geom_text_wordcloud() +
  theme_minimal()

encomium_cloud
```

That's underwhelming. Let's customize it a bit:
```{r wordcloud-pro}
ggplot(data = encomium_top100, aes(label = group, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "star") +
  scale_size_area(max_size = 30) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  theme_minimal()
ggsave("figures\\star.png",width=10)
```

Cool! And you can facet wrap (for different reports, for example) and update other aesthetics. See more here: https://cran.r-project.org/web/packages/ggwordcloud/vignettes/ggwordcloud.html

### Sentiment analysis

In the following, we will use Sentida to find sentiment scores of the entire text and individual words. Since an encomium is made to honour someone, we should expect the general sentiment to be quite positive.


```{r sentida}
# We will start by finding the total and mean sentiment scores for the encomium

encomium_total_sent <- sentida(iconv(encomium_no_numeric,"latin1", "UTF-8") , output = "total") 
encomium_mean_sent <- sentida(iconv(encomium_no_numeric,"latin1", "UTF-8") , output = "mean") 

encomium_total_sent
encomium_mean_sent

```
The mean tells us that the encomium in general is slightly more positive than negative. This was expected due to the genre demanding praise of someone, though it is surprising that the sentiment value isn't more positive than 0.22. Let's try finding the sentiment value of each individual word.


Let's find some counts (by sentiment ranking):
```{r count-sentida}
# We will start by making a data frame with the sentiment for each word in our 'encomium_no_numeric' tibble. 
# Firstly, we need a vector:

encomium_word_list <- encomium_no_numeric$group


# Now we can make a data frame that has the total sentiment value for each word using Sentida (lines 209-212 were partly generated by ChatGPT-5.5)

encomium_word_sent <- do.call(rbind,lapply(encomium_word_list,function(o) {
  x <- sentida(iconv(o,"latin1","UTF-8"),output="total")
  data.frame(ord=o,total=x)
}))


# Let's try plotting this with ggplot. We will order them based on their total sentiment value: 

encomium_word_sent %>% 
  ggplot(aes(x = reorder(ord, total), y = total)) +
  geom_col() +
  coord_flip() +
  theme_bw()

```

There are **a lot** of 0 value words that simply aren't in Sentida's lexicon. It is also very cluttered. Let us filter away the 0 value words, count only the top 20 of both positive and negative words, and color them depending on if they have a positive or negative value:

```{r sentida-20}
# Lines 229-245 were partly generated by ChatGPT-5.5
top20_sent_encomium_vis <- encomium_word_sent %>%
  filter(total != 0) %>%
  mutate(sentiment = ifelse(total > 0, "Positive", "Negative")) %>%
  group_by(sentiment) %>%
  slice_max(abs(total), n = 20) %>%
  ungroup() %>%
  arrange(total) %>%
  mutate(ord = factor(ord, levels = ord)) %>%
  ggplot(aes(x = ord, y = total, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(title = "Top 20 Positive and Negative Words in 'Anden Bog - Kong Knuts Liv og Gerninger'",
    x = NULL,
    y = "Total Sentiment Value") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y=element_blank())
top20_sent_encomium_vis
ggsave("figures/top20_sent_encomium.png", top20_sent_encomium_vis, width = 10, height = 5)

```

The positive words do in fact seem very positive. The top two negative words seem weird. Let's cross-check this by individually trying the words in sentida.

```{r sentida-20 more}
# So apparently "fandt" and "fandtes" are the absolute most negative words according to Sentida? That seems weird. We will try using Sentida to measure the total sentiment value of both these words individually:

sentida("fandt",output="total")
sentida("fandtes",output="total")

# Apparently yes, Sentida finds the past tenses of "finde" (Eng: "find") to be negative. Let's see if it finds "finde" or "fundet" (which is the past participle) negative too:

sentida("finde",output="total")
sentida("fundet",output="total")

# No. For some reason Sentida simply finds the past tense negative
```

It seems weird for "fandt" and "fandtes" to be that negative. It might be because being found is bad. Even then, being the absolute most negative word (ahead of "hade" and "ihjel) seems too much.
