---
title: "The Player Problem - The share of space portioned to different audio playback devices in Audio Magazine between 1973-1999"
author: "Marcus Fryd"
date: "2026-05-26"
output: 
  html_document:
    toc: true
    toc_float: true
---

### Packages used:
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)


library(tidyverse)
library(here)

# For text mining and plotting:

library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)

```



### Getting the Audio magazines and placing them in a tibble:

In the following, we will be loading the Audio magazines into R. The first step towards plotting them by year is combining all issues released in a given year into one PDF. To do this, we will be using 'pdf_combine()' from the pdftools package. 
After combining the PDFs, we need to load them into a tibble with their entire text. This is possible thanks to the 'pdf_text()' function from idem package.

```{r get-combine-documents}
# In the following, we will load the full text of the Audio Magazines. 

audio_files <- here("data/audio_1947_2000/used_issues") %>% # <- 'here()' defines the path 
  list.files(pattern="pdf$",full.names=T)  # <- 'list.files()' in this case lists all files with 'pdf' in the end (meaning all .pdf files) and their full name, meaning the entire path


# Now we will combine the PDFs by their year (lines 49-53 were partly generated by ChatGPT-5.5)

audio_year <- str_extract(basename(audio_files),"\\d{4}")
audio_files_by_year <- split(audio_files,audio_year)

lapply(names(audio_files_by_year), function(year) {
  pdf_combine(input = audio_files_by_year[[year]],
    output = paste0("/Users/fryga/Documents/rstudio projects/DAM_final_project/data/audio_by_year_1973_1999/", year,".pdf")
  )
})

# We have moved the output PDFs into a new folder under "/data". Below we will load in the full text of these new PDFs as a tibble (lines 58-61 were partly generated by ChatGPT-5.5)

audio_dir <- "data/audio_by_year_1973_1999"
audio_tbl <- tibble(
  file=list.files(audio_dir,pattern=".pdf$",full.names = TRUE)) %>%
  mutate(year=tools::file_path_sans_ext(basename(file)),
    text_clean=map(file,~paste(pdf_text(.x),collapse="\n"))) %>% 
  mutate(text_clean=str_split(text_clean, pattern = '\n')) %>% 
  unnest(text_clean) %>% 
  mutate(text_clean=str_trim(text_clean)) 
head(audio_tbl)

```



### Defining audio equipment, textmining, and plotting the findings:

Now that we have the full text in a tibble, we can start actually textmining. For our purpose, we will be using four "term groups", that is vectors containing (shortened) words for different audio playback devices. We will be looking for players for these four formats: Cassettes, CDs, records, and 8-tracks.

```{r mentions-plot}

# We will start by defining the term groups, after which we will plot the findings (lines 86-106 were partly generated by ChatGPT-5.5):

term_groups <- list(Turntable=c("turntable", "record pl", "turntables"),
  Cassette=c("cassette pl", "tapedeck", "cassette de","tapedecks", "cassette decks","compact cassette de","compact cassette pl","walkman","portable casset","tape pla"),
  CD=c("cdpl", "cd pl", "compact disc pl","cd-pl", "compact disk pl", "disc pl","cd-r pl","cd-rom pl", "cd-rw pl","cdchang","cd chang","compact disc chang","compact disk chang","cd-chang","discman","portable cd"),
  "8-track"=c("8-track pl","8-track d", "8 track pl","8-track tape pl","8 track d", "8 track tape pl","4-channel tape d","4 channel tape d","4-channel cartridge d","4 channel cartridge d","4-channel cartridge tape d","4 channel cartridge tape d")
)


audio_counts <- audio_tbl %>%
  bind_cols(
    map_dfc(term_groups, function(synonyms) {
      rowSums(
        sapply(synonyms, function(term) {
          str_count(audio_tbl$text_clean, regex(term, ignore_case = TRUE))
        }),
        na.rm = TRUE
      )
    }) %>%
      set_names(names(term_groups))
  )
audio_counts


yearly_counts <- audio_counts %>%
  group_by(year) %>%
  summarise(
  across(all_of(names(term_groups)), \(x) sum(x, na.rm = TRUE))
) %>%
  pivot_longer(-year, names_to = "Format", values_to = "Count")

yearly_counts$year <- as.numeric(yearly_counts$year)
yearly_counts

audio_graph <- yearly_counts %>% 
  group_by(Format) %>% 
  ggplot() +
  geom_line(aes(x=year,y=Count,color=Format,group=Format)) +
  scale_x_continuous(breaks=seq(1973,1999,by=1))+
  labs(title = "Total mentions of different home audio devices in Audio Magazine 1973-1999",
       x = "Year",
       y = "Total Mentions in Audio",
       color = "Category")+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(color = "grey20", angle = 45,
                                     hjust = 0.5, vjust = 0.5),
          axis.text.y = element_text(color = "grey20"),
          text = element_text(size = 11),
          plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
audio_graph
ggsave("figures\\audio_mentions_graph.png", plot = audio_graph, width = 10, height = 5, bg = "white")

```

### Getting and plotting Recording Industry Association of America data:

We can now load in the data from our "format_revenue" table with the 'read_csv2' function from the tidyverse package. This can then easily be plotted in a similar fashion to how we plotted the data from our textmining above:

```{r riaa-plot}

# We start by loading in the tidied up dataset as a tibble:

format_revenue <- read_csv2("data/format_revenue.csv")
head(format_revenue)

# Before continuing with plotting the data, we have to ensure the graph shows us standard form of number and not the scientific notation, which becomes the default when dealing with larger number. To ensure this, we execute the following line (credits to Sacha Epskamp on Stack Overflow)

options(scipen = 999)

# We can now plot our data with the 'geom_line' function of ggplot2. 

format_revenue_plot<- format_revenue %>% 
  filter(Year<"2000") %>% 
  ggplot()+
  geom_line(mapping=aes(x=Year, y=Revenue,
                        color=Format,))+
  scale_x_continuous(breaks=seq(1973,1999,by=1))+
  labs(title="Revenue of physical music formats 1973-1999",
       caption = "Data from the RIAA compiled by Throwback Thursday (data.world)",
    x="Year",
    y="Revenue in USD (not adjusted for inflation)",
    color="Format")+
  theme_bw()+
  theme(legend.position = "bottom", 
        axis.text.x = element_text(color = "grey20", angle = 45,
                                     hjust = 0.5, vjust = 0.5),
          axis.text.y = element_text(color = "grey20"),
          text = element_text(size = 11),
          plot.title = element_text(hjust = 0.5))
format_revenue_plot
ggsave("figures\\formats_revenue_graph.png", plot = format_revenue_plot, width = 10, height = 5, bg = "white")
```

### Investigating most common words in Audio Magazine before and after 1987

In the following chunk, we will being investigating what the most common words were in Audio Magazine after and including 1987, since we saw a dip in the mentions of CD players. We will similarly be investigating what the most common words were before 1987 afterwards

```{r count-words-1987}

# We will start by tokenizing our audio_tbl filtered to only contain the years 1987 to 1999:

audio_tokenize_87 <- audio_tbl %>% 
  filter(year>=1987) %>% 
  unnest_tokens(word, text_clean)
audio_tokenize_87


# In the following, we will load in an English stopword list.
## Afterwards, we will use the 'anti_join()' function to remove the stopwords from the tokenized tibble:

stop_words_eng <- read_csv("data/english_stopwords.txt")
  
audio_stop_87 <- audio_tokenize_87 %>% 
  anti_join(stop_words_eng)


# Below, we will filter all numbers, since those are not particularly interesting:

audio_no_numeric_87 <- audio_stop_87 %>% 
  filter(is.na(as.numeric(word))) 


# We will now count all words and sort them from most to fewest mentions:

audio_word_count_87 <- audio_no_numeric_87 %>% 
  count(word) %>% 
  group_by(word) %>%
  summarise(n = sum(as.numeric(n)), .groups = "drop") %>% 
  arrange(-n) 
audio_word_count_87

```

Since the process is the same as above, there are no explanations as to what is being done.

```{r count-words-1973}

audio_tokenize_73 <- audio_tbl %>% 
  filter(year<1987) %>% 
  unnest_tokens(word, text_clean)
audio_tokenize_73


audio_stop_73 <- audio_tokenize_73 %>% 
  anti_join(stop_words_eng)


audio_no_numeric_73 <- audio_stop_73 %>% 
  filter(is.na(as.numeric(word))) 


audio_word_count_73 <- audio_no_numeric_73 %>% 
  count(word) %>% 
  group_by(word) %>%
  summarise(n = sum(as.numeric(n)), .groups = "drop") %>% 
  arrange(-n) 
audio_word_count_73

```

### Wordclouds of the top 100 most common words in Audio Magazine before and after 1987

We will now make two wordclouds over the top 100 most words in Audio Magazine before and after 1987. We will not do a wordcloud over all the words, since that would be far too large to analyze anything interesting. 
To do this, we will have to firstly make a top 100 tibble, which is easily done with the 'head()' function. We will start with after 1987:

```{r top-words-1987}

# By writing 100 in the 'head()' function, we can view only the top 100, so it is only a matter of assigning the function a new name:

audio_top100_87 <- audio_word_count_87 %>% 
  head(100)
audio_top100_87

```

Now, onto before 1987:

```{r top-words-1973}

audio_top100_73 <- audio_word_count_73 %>% 
  head(100)
audio_top100_73

```

We are now able to make wordclouds using the 'geom_text_wordcloud_area' function under a ggplot pipeline. Let us begin with 1987-1999:

```{r wordcloud-1987}

# We will determine frequency by color: The more frequently a word appeares (i.e., the higher an "n" value), the redder. Afterwards, less frequent words (a lower "n" value) turn purple, then blue. The least frequent of the top 100 most frequent words will be green.

audio_t100_word_cloud_87 <- ggplot(data = audio_top100_87, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n)) +
  scale_size_area(max_size = 30) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  labs(title="Top 100 most frequent words in Audio Magazine after and including 1987")+
  theme_minimal()+
  theme(text = element_text(size = 11),
        plot.title = element_text(hjust = 0.5))
audio_t100_word_cloud_87
ggsave("figures\\audio_t100_word_cloud_87.png", plot = audio_t100_word_cloud_87, bg = "white")

```

And now before 1987 (same as above):

```{r wordcloud-1973}

audio_t100_word_cloud_73 <- ggplot(data = audio_top100_73, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n)) +
  scale_size_area(max_size = 30) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  labs(title="Top 100 most frequent words in Audio Magazine before 1987")+
  theme_minimal()+
  theme(text = element_text(size = 11),
        plot.title = element_text(hjust = 0.5))
audio_t100_word_cloud_73
ggsave("figures\\audio_t100_word_cloud_73.png", plot = audio_t100_word_cloud_73, bg = "white")

```

It is very quite difficult to see which words are unique to the period 1987-1999. We can use the 'anti_join' function on our top 100 objects to remove the words not unique to the period after 1987 and plot that as a wordcloud:

```{r anti-join}

# We start by anti-joining the two top 100 objects from above

audio_73_87_anti <- audio_top100_87 %>% 
  anti_join(audio_top100_73,
            by = "word")
audio_73_87_anti

# We can now use the same wordcloud code from before to plot this:

audio_t100_word_cloud_anti <- ggplot(data = audio_73_87_anti, aes(label = word, size = n)) +
  geom_text_wordcloud(aes(color = n)) +
  scale_size_area(max_size = 15) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  labs(title="Most frequent words in Audio Magazine that were unique to the period 1987-1999")+
  theme_minimal()+
  theme(text = element_text(size = 11),
        plot.title = element_text(hjust = 0.5))
audio_t100_word_cloud_anti
ggsave("figures\\audio_t100_word_cloud_anti.png", plot = audio_t100_word_cloud_anti,width = 7, height = 4, bg = "white")

```
